Verification of concurrent design patterns with data

### 2019-02-19 2018-Arch-pNets/arch-pNets.tex
arch-pNets paper: Some improvements in - the discussion of alphas (fresh names of the SVs in the encoding) - verification discussion at end of Sect 4 NB: Added the Boolean variables to the alphas in the encoding.

\newcommand{\addspace}[1]{#1 }
\newcommand{\nopri}[1][]{\ensuremath{\mathit{pNet}%
    \ifx#1\relax\else(#1)\fi}}
\newcommand{\maxprog}[1][]{\ensuremath{\mathit{pNet}_\mu%
    \ifx#1\relax\else(#1)\fi}}

@@

  \label{Example:TimerOT3}

@@

We obtain the
synchronisation vector SV0 shown in \fig{FailureTimer:pNet}, where $b0$ and
$b1$ are the Boolean values associated to the actions encoding the
ports {\PortFail} and {\NameCtrl.\PortFail}.
The guard $b_1 \Rightarrow b_0$ encodes the causal
relation between ports {\NameCtrl.\PortFail} and {\PortFail}.  Notice
that, in our encoding, all three ports are present in the
synchronisation vector.  \Fig{FailureTimer:pNet} shows
the four synchronisation vectors SV0\ndash SV3 corresponding to the
connectors in \fig{schema:ArchFailure:BIP} and an additional vector
SV4, synchronising the {\init} transitions of the two pLTSs.

\addSB{In the general encoding, the global actions of the
  synchronisation vectors should be fresh action names.  However, for
  the properties considered in our case study, we only need to exhibit
  the actions of the holes.  Furthermore, each synchronisation vector
  involves at most one such action.  This allows us to reuse them as
  the global actions of the synchronisation vectors, simplifying the
  translation of properties to be verified (\cf \app{properties}).}
\noteLH{ce paragraphe sonne vraiment note de rapport de recherche, tu arrives a lui donner un peu d'abstraction/de recul?}

In the general case, the encoding relies on the causal semantics of the
algebra of BIP connectors \cite{BliSif10-causal-fmsd}.  Disregarding
the variables and data transfer, the Algebra of Connectors $\ac(P)$
\cite{BliSif07-acp-emsoft} provides a syntactic notation for the BIP
connectors.  The causal semantics of the connectors, given in terms of
the Algebra of Causal Interaction Trees $\ct(P)$, explicitates the
causal dependencies discussed above through an encoding mapping
$\tau: \ac(P) \rightarrow \ct(P)$.  Another mapping $R: \ct(P)
\rightarrow \cru(P)$ encodes causal interaction trees into systems of
causal rules, which are Boolean implications similar to the ones in
the example above.  We briefly summarise the key definitions and
results in \app{algebras}.  The $\ac(P)$, $\ct(P)$ and $\cru(P)$
representations of the connectors in \fig{schema:ArchFailure:BIP} are
shown in \tab{connectors} (elements shown in red can be removed for
simplification as described in the example above).

@@

\noteEM{Dans le paragraph precedent, tu resume le cas des connecteurs sans data, dans les prochains, tu formalise avec data, mais pour le lecteur ``casual'', le changement n'est pas evident... }
Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then, \addSB{the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
  and the encoding of $\Gamma$ is} $\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  \alpha_p
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%% \noteSB{Ought to group ports by components in
%%   $\Tuple{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}^{p \in P_x}$...
%%   \newline
%%   Will probably restrict interactions to one port per component
%%   throughout the paper to simplify.
%% }
%
with $b_p$ a fresh Boolean variable, $\alpha_p$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that preceeds it
all occurrences of $p$ by $b_p$.

@@

\paragraph{Encoding of the maximal progress priority model}
We only present the key idea, which consists in
introducing an additional Boolean variable, for each port, for which we
have introduced one in $\nopri[A, \partition]$.  Intuitively, 
the Boolean
variables introduced for the encoding of interaction models
determine whether it is the original transition
labeled by the port that is executed (\true), or rather the corresponding
self-loop, introduced by the encoding (\false).  The new variables
determine whether \emph{there is an original transition labelled by
  $p$ that could be executed from the same state}, \ie[,] with $b_p$
the variable introduced above for the encoding without maximal
progress, $q \goesto{p(\true, b_p)} q'$ iff $\exists q'': q
\goesto{p(b_p)} q''$ with $b_p = \true$ in $\nopri[A, \partition]$.
Then, in the SV guard, we have to check whether \emph{all ports
  leading to $p$ in the causal tree $\tau(x)$ can be fired} (see
\app{transformations}).  If so, then $p$ must be fired, \ie
$p(\false)$ must be blocked.
%% This is pretty much straightforward, but
%% care must be taken when doing this in the general case, since causal
%% interaction trees can get pretty tricky due to $\oplus$ (see
%% \app{trees}).}
%****************************************************************
%****************************************************************

@@

\label{secn:arch:verif}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{TimerOADetailed}
  \caption{The open Automaton of the Failure Monitor Architecture}
  \label{schema:ArchFailure:OA}
\end{figure}

In \cite{HMZ-FORTE2016QBMZ-AVOCS18}, we have described an algorithm
and a tool that 
computes the open automaton encoding the semantics of an open
pNet. This tool computes open transitions from the combinations of all
possible behaviours and synchronisation vectors of the pNet, then makes
use of an SMT engine to check satisfiability of the open transition
predicates, and minimize the size of the resulting automaton.

@@

In Figure~\ref{schema:ArchFailure:OA} we show the full open automaton
obtained from our pNet. For reasons of space, we have not shown the
detail of the open transition, but only the assignments of state
variables, some usefull parts of the predicates, and the resulting
action; you can find the full details in 
Appendix \ref{secn:full-results}. As an example, $OT_3$ is the one
explained in Example~\ref{Example:TimerOT3} page~\pageref{Example:TimerOT3}

This automaton has 12 transitions, including those encoding various
possible firing of some interactions (e.g. OT 2 and 3 for
$fail$). At first sight, it appears the rough behaviour is
reasonable, although a detailed study shows that the property stated
in page~\pageref{property:reset} does not hold, because of the
$fail$ loops on states $11$ and $12$. This is because we did not use
the maximal progress assumption here. If we do, we get the corrected
behaviour in Fig~\ref{schema:ArchFailure:OA-MaxProgress}
(\app{full-results}), in which $OT_7$ has 
disappeared, and $OT_2$ and $OT_{10}$ have been restricted to
$b0=false$. This one verifies all the properties listed in
\app{properties}.

@@

Now we list the full set of OTs, in latex form, for better
readability. This comes as a complement to the open automaton in
Fig. \ref{schema:ArchFailure:OA-annex}, in which we only have a partial view
of the OTs details.

\begin{figure}[t]
  \centering
  \includegraphics[width=\columnwidth]{TimerOAFullDetailed}
  \caption{The open Automaton of the Failure Monitor Architecture}
  \label{schema:ArchFailure:OA-annex}
\end{figure}

@@

  In Fig. \ref{schema:ArchFailure:OA-annex}, you can see that some
  properties from appendix \app{properties} do not hold.

@@

  The new OA is shown in figure \ref{schema:ArchFailure:OA-MaxProgress}.

  \begin{figure}[t]
  \centering
  \includegraphics[width=\columnwidth]{TimerOAMaxProgress}
  \caption{The open Automaton of the Failure Monitor Architecture,
    with Maximal Progress}
  \label{schema:ArchFailure:OA-MaxProgress}
\end{figure}

@@

\bigskip
Fig. \ref{schema:ArchFailure:OA-TimerV4X2} shows the open automata of
this composed architecture.
Note that e..g. the open transition labelled ``$fail(*,tt)$'' stands for
``$fail(b01,b02)$'', where  $b01'$ can be arbitrarily true or
false. As in the previous case, this is because we have not use here
the Max Progress hypothesis, otherwise all the ``*'' in the figure
should be replaced by ``$ff$''.

  \begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{TimerV4X2}
  \caption{The open Automaton of the composed system}
  \label{schema:ArchFailure:OA-TimerV4X2}
\end{figure}

>>>>

\newcommand{\addspace}[1]{#1 }
\newcommand{\nopri}[1][]{\ensuremath{\mathit{enc}%
    \ifx#1\relax\else(#1)\fi}}
\newcommand{\maxprog}[1][]{\ensuremath{\mathit{enc}_\mu%
    \ifx#1\relax\else(#1)\fi}}

@@

  \label{ex:TimerOT3}

@@

We obtain the
synchronisation vector SV0 shown in \fig{FailureTimer:pNet}, where $b0$ and
$b1$ are the Boolean values associated to the actions encoding the
ports {\PortFail} and {\NameCtrl.\PortFail}.
The guard $b1 \Rightarrow b0$ encodes the causal
relation between these ports.  Notice
that %% , in our encoding,
all three ports are present in the
synchronisation vector.  \Fig{FailureTimer:pNet} shows
the four synchronisation vectors SV0\ndash SV3 corresponding to the
connectors in \fig{schema:ArchFailure:BIP} and an additional vector
SV4, synchronising the {\init} transitions of the two pLTSs.

In the general case, the encoding relies on the causal semantics of the
algebra of BIP connectors \cite{BliSif10-causal-fmsd}.  Disregarding
the variables and data transfer, the Algebra of Connectors $\ac(P)$
\cite{BliSif07-acp-emsoft} provides a syntactic notation for the BIP
connectors.  The causal semantics of the connectors, given in terms of
the Algebra of Causal Interaction Trees $\ct(P)$, explicitates the
causal dependencies through an encoding mapping
$\tau: \ac(P) \rightarrow \ct(P)$.  Another mapping $R: \ct(P)
\rightarrow \cru(P)$ encodes causal interaction trees into systems of
causal rules, which are Boolean implications similar to the ones in
the example above.  We briefly summarise the key definitions and
results in \app{algebras}.  The $\ac(P)$, $\ct(P)$ and $\cru(P)$
representations of the connectors in \fig{schema:ArchFailure:BIP} are
shown in \tab{connectors} (elements shown in red can be removed for
simplification as described in the example above).

@@

\noteEM{Dans le paragraph precedent, tu resume le cas des connecteurs sans data, dans les prochains, tu formalise avec data, mais pour le lecteur ``casual'', le changement n'est pas evident... }
Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignement
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then, \addSB{the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
  and the encoding of $\Gamma$ is} $\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, where $\nopri[x,g,u]$ is the synchronisation vector defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  {\color{blue}\alpha \bigl(b_p^{p \in P_x}\bigr)}
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%
with $b_p^{p \in P_x}$ fresh Boolean variables, $\alpha$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that preceeds it
all occurrences of all $p$ by corresponding variables $b_p$.

\addSB{%
%
For the sake of clarity, we simplify the encoding in
\fig{FailureTimer:pNet} by reusing the port names of the original
architecture instead of fresh names $\alpha$.  This is made
possible by the fact that each synchronisation vector involves at
most one action of interest (see the properties in
\app{properties}).
%% In the general encoding, the global actions of the
%% synchronisation vectors should be fresh action names.
%% However,
%% for the properties considered in our case study, we only need to
%% exhibit the actions of the holes.  Furthermore, each
%% synchronisation vector involves at most one such action.
%% This
%% allows us to reuse them as the global actions of the
%% synchronisation vectors, simplifying the translation of
%% properties to be verified (\cf \app{properties}).
%
}\noteLH{ce
  paragraphe sonne vraiment note de rapport de recherche, tu
  arrives a lui donner un peu d'abstraction/de recul?}\noteSB{Est-ce que c'est mieux comme \c{c}a ?}

@@

\paragraph{Encoding of the maximal progress}
We only present the key idea, which consists in
introducing an additional Boolean variable, for each port, for which we
have introduced one in $\nopri[A, \partition]$.  Intuitively, 
the Boolean
variables introduced for the encoding of interaction models
determine whether it is the original transition
labeled by the port that is executed (\true), or rather the corresponding
self-loop, introduced by the encoding (\false).  The new variables
determine whether \emph{there is an original transition labelled by
  $p$ that could be executed from the same state}, \ie[,] with $b_p$
the variable introduced above for the encoding without maximal
progress, $q \goesto{p(\true, b_p)} q'$ iff $\exists q'': q
\goesto{p(b_p)} q''$ with $b_p = \true$ in $\nopri[A, \partition]$.
In the SV guard, we have to check whether \emph{all ports
  leading to $p$ in the causal tree $\tau(x)$ can be fired} (see
\app{transformations}).  If so, then $p$ must be fired, \ie
$p(\false)$ must be blocked.
%% This is pretty much straightforward, but
%% care must be taken when doing this in the general case, since causal
%% interaction trees can get pretty tricky due to $\oplus$ (see
%% \app{trees}).}
%****************************************************************
%****************************************************************

@@

%% \label{secn:arch:verif}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{TimerOADetailed}
  \caption{The open Automaton of the Failure Monitor Architecture}
  \label{fig:ArchFailure:OA}
\end{figure}

\bigskip\noteSB{J'ai repris ce paragraphe}
The above encoding provides a mechanism for the symbolic verification
of architectures using our existing tool
\cite{HMZ-FORTE2016QBMZ-AVOCS18} to compute the open automaton
semantics of an open pNet. This tool computes open transitions from
pLTS behaviours and synchronisation vectors of the pNet, then uses an
SMT engine to check satisfiability of their predicates, minimizing the
size of the resulting automaton.

@@

In \fig{ArchFailure:OA} we show the full open automaton
obtained from the pNet in \fig{FailureTimer:pNet}.  Due to space limitations, we do not show the
details of the open transitions, but only the assignments of state
variables, some useful parts of the predicates, and the resulting
action; full details can be found in 
\app{full-results}.
%
\noteSB{Removed ``As an example...'', since that \ex{TimerOT3} was
  simplified and does not correspond to the real OT.}
%% As an example, $OT_3$ is the one
%% explained in \ex{TimerOT3} (page~\pageref{ex:TimerOT3}).

This automaton has 12 transitions, including those encoding various
possible firing of some interactions (e.g. OT 2 and 3 for
{\PortFail}). At first sight, it appears the rough behaviour is
reasonable, although a detailed study shows that the property stated
in page~\pageref{property:reset} does not hold, because of the
{\PortFail} loops on states $11$ and $12$. This is because we did not
use the maximal progress assumption here. If we do, we get the
corrected behaviour in \fig{ArchFailure:OA-MaxProgress}
(\app{full-results}), in which $OT_7$ has disappeared, and $OT_2$ and
$OT_{10}$ have been restricted to $b0=\false$. This one verifies all
the properties listed in \app{properties}.

@@

Now we list the full set of OTs, in latex form, for better
readability. This comes as a complement to the open automaton in
\fig{ArchFailure:OA-annex}, in which we only have a partial view
of the OTs details.

\begin{figure}[t]
  \centering
  \includegraphics[width=\columnwidth]{TimerOAFullDetailed}
  \caption{The open Automaton of the Failure Monitor Architecture}
  \label{fig:ArchFailure:OA-annex}
\end{figure}

@@

  In \fig{ArchFailure:OA-annex}, you can see that some
  properties from appendix \app{properties} do not hold.

@@

  The new OA is shown in \fig{ArchFailure:OA-MaxProgress}.

  \begin{figure}[t]
  \centering
  \includegraphics[width=\columnwidth]{TimerOAMaxProgress}
  \caption{The open Automaton of the Failure Monitor Architecture,
    with Maximal Progress}
  \label{fig:ArchFailure:OA-MaxProgress}
\end{figure}

@@

\bigskip
\Fig{ArchFailure:OA-TimerV4X2} shows the open automata of
this composed architecture.
Note that e..g. the open transition labelled ``$fail(*,tt)$'' stands for
``$fail(b01,b02)$'', where  $b01'$ can be arbitrarily true or
false. As in the previous case, this is because we have not use here
the Max Progress hypothesis, otherwise all the ``*'' in the figure
should be replaced by ``$ff$''.

  \begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{TimerV4X2}
  \caption{The open Automaton of the composed system}
  \label{fig:ArchFailure:OA-TimerV4X2}
\end{figure}
